Teaching modeling in introductory statistics: A comparison of formula and tidyverse syntaxes

### 2022-05-12 PaperDraft.Rmd
revised according to reviewer comments

---
title: "Teaching modeling in introductory statistics: A comparison of formula and tidyverse syntaxes"

# to produce blinded version set to 1
blinded: 0

authors:
- name: Amelia McNamara ^[amelia.mcnamara@stthomas.edu]
  affiliation: |
    | Department of Computer & Information Sciences, University of St Thomas
  email: amelia.mcnamara@stthomas.edu
keywords: |
  R language, instruction, data science, statistical computing

abstract: >
  This paper reports on an experiment run in a pair of introductory statistics labs, attempting to determine which of two R syntaxes was better for introductory teaching and learning: formula or tidyverse. One lab was conducted fully in the formula syntax, the other in tidyverse. Analysis of incidental data from YouTube and RStudio Cloud show interesting distinctions. The formula section appeared to watch a larger proportion of pre-lab YouTube videos, but spend less time computing on RStudio Cloud. Conversely, the tidyverse section watched a smaller proportion of the videos and spent more time on RStudio Cloud. Analysis of lab materials showed that tidyverse labs tended to be slightly longer (in terms of lines in the provided RMarkdown materials, as well as minutes of the associated YouTube videos), and the tidyverse labs exposed students to more distinct R functions. However, both labs relied on a quite small vocabulary of consistent functions. Analysis of pre- and post-survey data show no differences between the two labs, so students appeared to have a positive experience regardless of section. This work provides additional evidence for instructors looking to choose between syntaxes for introductory statistics teaching. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    keep_tex: yes
    includes:
      in_header: header.tex
header-includes:
-  \usepackage{amsmath}
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{caption}
- \usepackage{subcaption}
- \usepackage{xcolor,soul}
- \usepackage{xspace}
editor_options: 
  chunk_output_type: console
---

@@

lablines <- read_csv(here("data", "anonymized", "lablines.csv"),
  col_types = cols(
    file = col_character(),
    lines = col_double(),
    section = col_character()
  )
) %>%
  mutate(section = if_else(section == "tidy", "tidyverse", section))

@@

In R, the development of less-complex code for students has been under consideration for even longer. R offers non-standard evaluation, which allows package authors to create new 'syntax' for their packages [@morandatetal2012]. In human language, syntax is the set of rules for how words and sentences should be structured. If you use the wrong syntax in human language, people will probably still understand you, but they will be able to hear there is something wrong with how you structured your speech or writing. Syntax in programming languages is even more formal-- it governs what code will execute, run, or compile correctly. Using the wrong syntax means getting an error from the language.

@@

There is some agreement about pedagogical decisions for teaching R. In particular, most educators agree that in order to reduce cognitive load, instructors should only teach one syntax, and to be as consistent as possible about that syntax [@mcnamaraetal2021]. There is also some agreement base R syntax is not the appropriate choice for introductory statistics, but there is widespread disagreement on whether the formula syntax or `tidyverse` syntax is better for novices.

While there are strongly-held opinions on which syntax should be taught [@pruimetal2017; @cetinkaya-rundeletal2021], there is relatively little empirical evidence to support these opinions. In the realm of computer science, empirical studies by Andreas Stefik, et al have shown significant differences in the intuitiveness of languages, as well as error rates, based on language design choices [@stefiketal2011; @stefiksiebert2013]. Thus, it seems likely there are language choices that could make data science programming easier (or harder) for users, particularly novices.

@@

The remainder of this paper is organized into three sections. Section \ref{sec:methods} describes the setup of the study, the participants (\ref{sec:participants}) and their experience (\ref{sec:priorexp}), and the content of the course under investigation (\ref{sec:materials}). Section \ref{sec:results} contains results of the analysis, including a comparison of material lengths between the sections (\ref{sec:ll}), the number of unique functions shown in each section (\ref{sec:numfunc}), results from the pre- and post-survey (\ref{sec:prepost}), and analysis of YouTube (\ref{sec:yt}) and RStudio Cloud (\ref{sec:rstudio}) data. Finally, Section \ref{sec:discussion} discusses the results and opportunities for future study.

All materials used for this study are available on GitHub and are Creative Commons licensed, so they can be used or remixed by anyone who wants to use them. All code and anonymized data from this paper is also available on GitHub, for reproducibility. Data analysis was performed in R, and the paper is written in RMarkdown. The categorical color palette was chosen using Colorgorical [@gramazioetal2017], and colors for the Likert scale plot are from ColorBrewer [@harrowerbrewer2003]. Example data used throughout the paper is from `palmerpenguins` [@horstetal2020]. Packages used for the formula section were `mosaic` and `ggformula` (now loaded automatically with `mosaic`), for the tidyverse section the `tidyverse` and `infer` packages [@pruimetal2017; @kaplanpruim2020; @wickhametal2019; @brayetal2021].

@@

The author ran a pilot study in her introductory statistics labs. This study was run twice, once in the Spring 2020 semester and once in the Fall 2020 semester. The disruption of COVID-19 to the Spring 2020 semester made the resulting data unusable, so this paper focuses on just Fall 2020 data.

Data was collected from YouTube analytics for watch times, from RStudio Cloud for aggregated compute time, and from pre- and post-surveys of students. Participants for the pre- and post-survey were recruited from this pool after Institutional Research Board ethics review.

@@

Participants in the study were students enrolled in an introductory statistics course at a mid-sized private university in the upper Midwest. At this university, statistics students enroll in a lecture (approximately 60-90 students per section), which is broken into several smaller lab sections for hands-on work in statistical software. Lecture and lab sections are taught by different instructors, and the lab sections associated with a particular lecture often use different software. For example, one lab may use Minitab while the other two use Excel. However, every lab section (no matter what lecture it is associated with, or what software is used) does the same set of standardized assignments. This structure provides a consistent basis for comparison.

@@

Each week, the lab instructor prepared a "pre-lab" document in RMarkdown. The pre-lab covered the topics necessary to complete the standardized lab assignment done by all students across lab sections. Pre-lab documents included text explanations of statistical and R programming concepts, sample code, and blanks (both in the code and the text) for students to fill in as they worked. The instructor recorded YouTube videos of herself working through the pre-lab documents for each lab, and posted them in advance. Students were told to watch the pre-lab video and work through the RMarkdown document on their own time, then come to synchronous class to ask questions and get help starting on the real lab assignment.

@@

One obvious question arising when considering the comparison of the two syntaxes is whether students performed better in one section or another. The IRB for this study did not cover examining student work (an obvious place for improved further research), so we cannot look at student outcomes on a per-assignment basis. However, running a randomization test for a difference in overall mean lab grades showed no significant difference between the two sections. While they may have been interesting differences in grades depending on the topic of the lab, we at least know these differences averaged out in the end.

@@

@@

```{r}
labsfortesting <- lablines %>%
  drop_na(file) %>%
  pivot_wider(week, names_from = section, values_from = lines) %>%
  mutate(diff = tidyverse - formula)
library(infer)
set.seed(11082021)
labboot <- labsfortesting %>%
  specify(response = diff) %>%
  generate(reps = 5000, type = "bootstrap") %>%
  calculate(stat = "mean") %>%
  get_ci()
```

The length of the pre-lab RMarkdown documents can be measured using lines. Figure \ref{fig:prelablength} shows the number of lines of code for each section's pre-lab document, per week.

It indicates RMarkdown documents for the `tidyverse` section tended to be longer. We can compute a difference in lab lengths for each week, and compute the mean difference, which is `r round(mean(labsfortesting$diff))` lines. Because we only have 12 labs worth of data, we used a bootstrap procedure to generate a confidence interval for the mean of the differences. The 95\% interval is (`r round(labboot$lower_ci)`, `r round(labboot$upper_ci)`), which indicates labs for the `tidyverse` section were longer, but only by a few lines.

A slightly longer length for these labs makes sense, because `tidyverse` code is characterized by multiple short lines strung together into a pipeline with `%>%`, while the formula syntax typically has single function calls, sometimes with more arguments.

@@

youtube_videos %>%
  ggplot(aes(x = week, y = tot_min, fill = section)) +
  geom_col(position = "stack", color = "white") +
  scale_x_continuous(breaks = seq(from = 1, to = 15, by = 2)) +
  facet_wrap(~section, nrow = 2) +
  ylab("Total length of pre-lab \n videos (minutes)") +
  xlab("Week of semester") +
  theme(
    legend.position = "none"
  )
```

```{r}
youtube_fortesting <- youtube_weeks %>%
  select(week, section, tot_min) %>%
  pivot_wider(week, names_from = section, values_from = tot_min) %>%
  mutate(diff = tidyverse - formula, 
         avglength = (tidyverse + formula)/2, 
         perc_diff = abs(tidyverse - formula)/avglength)

ytboot <- youtube_fortesting %>%
  specify(response = diff) %>%
  generate(reps = 5000, type = "bootstrap") %>%
  calculate(stat = "mean") %>%
  get_ci()
detach("package:infer")
```

@@

Again, we can compute a pairwise difference in total video length (adding together multiple videos in weeks that had them), and compute the mean of that difference. That difference is `r round(mean(youtube_fortesting$diff))` minutes (`tidyverse` videos being longer). We can then compute a 95\% bootstrap confidence interval for the difference, (`r round(ytboot$lower_ci, digits=2)`, `r round(ytboot$upper_ci)`). Again, it appears `tidyverse videos are longer, although just by a few minutes.

@@

A similar discrepancy can be seen in week 10, where the formula section's RMarkdown document was `pull(filter(lablines, week == 10, section == "formula"), lines)` lines long, and the videos totaled `r round(filter(youtube_weeks, week == 10, section == "formula") %>% pull(tot_min))` minutes. That same week the `tidyverse` RMarkdown document was `pull(filter(lablines, week == 10, section == "tidyverse"), lines)` lines long, and the videos totaled `r round(filter(youtube_weeks, week == 10, section == "tidyverse") %>% pull(tot_min))` minutes.

@@

These discrepancies made it take longer to explain code in the `tidyverse` section. Comparisons of RMarkdown document length and YouTube video length, as well as the corresponding reasons for those discrepancies are the first hint of the computing time results to come in Section \ref{sec:rstudio}.

@@

Of course, while teaching a real class, an instructor often has to ad-lib at least a little. So, it is also interesting to consider the number of functions actually shown throughout the course of the semester. To do this, we can consider the functions shown in the filled-in version of pre-lab documents the instructor ended up with after recording the associated instructional video.

@@

The functions both sections of students saw included helper functions like `library()`, `set.seed()`, and `set()` (a function in the knitr options included in the top of each RMarkdown document), statistics like `mean()`, `sd()`, and `cor()`, and modeling-related functions like `aov()`, `lm()`, `summary()` and `predict()`.

@@

Students in the formula section saw `r nrow(formulafunctions)` functions unique from the set both sections saw, while the `tidyverse` section saw `r nrow(tidyfunctions)` unique functions. It makes sense the number of unique functions in the `tidyverse` section would be slightly larger. One reason is the `ggplot2` helper functions `ggplot()` and `aes()`.

Students in both sections saw how to make a barchart, boxplot, histogram, and scatterplot, but in the formula section they used standalone functions like `gf_boxplot()` whereas in the `tidyverse` section they needed to start with `ggplot` and add on a `geom_` function like `geom_boxplot()`, while specifying the `aes`thetic values somewhere.

Similarly, both sections saw several common summary statistics, but in the formula section they used the function (e.g. `mean()`) on its own, whereas in the `tidyverse` section summary functions needed to be wrapped within `summarize()`. Students in the `tidyverse` section also saw slightly more summary statistic functions, because one lab called for the five number summary.

@@

In the `tidyverse` section, the instructor chose to show two approaches. (Probably a bad pedagogical decision.) Both approaches are in \ref{tidy-summary1}, and both needed to include `drop_na()` to deal with missing values. Past those similarities, the approaches are divergent.

@@

The instructor should have chosen a single solution to present to students, but was faced with a dilemma. The first `tidyverse` approach is very verbose, but it follows nicely from other summary statistics students had already seen, just adding a few more functions like `min`, `max`, and `quantile`. The second solution is more concise, but it introduces the `pull` function, which was never used again in the course.

This brings up an important consideration when teaching coding-- how many times students will see the same function. Because there is some cognitive load associated with learning a new function, and repetition helps move information from working memory to long term memory, it is ideal for students to see each function at least twice [@lovettgreenhouse2000; @mcnamaraetal2021a]. When analyzing the number of functions shown in each section, we found there were `r dim(filter(allfunctions, n == 1, section == "formula"))[1]` functions shown only one time in the formula section, and `r dim(filter(allfunctions, n == 1, section == "tidyverse"))[1]` functions only shown once in the `tidyverse` section.

The practice of analyzing the number of functions shown over the course of the semester was eye-opening. It will provide valuable information for the instructor the next time she teaches the course, as she can attempt to remove functions only shown once, and ensure the cheatsheets better match what is actually shown throughout the semester.

@@

As discussed in \ref{sec:participants}, the number of students who completed both the pre- and post-surveys were low, so there is limited generalizability of the paired analysis.

@@

\begin{itemize}
\itemsep-3mm
```{r, results = 'asis'}
qs %>%
  select(Q11.x:Q16.y) %>%
  pivot_longer(cols = everything(), names_to = "old", values_to = "q") %>%
  select(-old) %>%
  mutate(q = paste("\\item", q, "\n")) %>%
  pull(q) %>%
  walk(cat)
```
\end{itemize}

In Figure \ref{fig:pre-post}, you can see a visualization of these Likert-scale questions, split by section.

```{r}
pre_sum <- prepostsurvey %>%
  select(Section.x, Q11.x:Q16.x) %>%
  pivot_longer(-Section.x,
    names_to = "skill",
    values_to = "feeling"
  )
pre_sum <- pre_sum %>%
  drop_na(feeling) %>%
  group_by(Section.x, skill, feeling) %>%
  summarize(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  select(-n) %>%
  mutate(type = "pre") %>%
  rename(Section = Section.x)

post_sum <- prepostsurvey %>%
  select(Section.y, Q11.y:Q16.y) %>%
  pivot_longer(-Section.y,
    names_to = "skill",
    values_to = "feeling"
  )
post_sum <- post_sum %>%
  drop_na(feeling) %>%
  group_by(Section.y, skill, feeling) %>%
  summarize(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  select(-n) %>%
  mutate(type = "post") %>%
  rename(Section = Section.y)

@@

```{r pre-post, fig.cap = "Pre and post responses to Likert-scale questions. Most questions show some level of improvement, such as the first question, `I am confident in my ability to make use of programming software to work with data.' but others show no change or even a decline in agreement.", out.width = "95%", fig.asp = 0.618}
likert_palette <- c("1 - strongly disagree" = "#a6611a", "2" = "#dfc27d", "3 - neutral" = "grey90", "4" = "#80cdc1", "5 - strongly agree" = "#018571")
ggplot(prepost_pos, aes(x = type, y = prop, fill = feeling)) +
  geom_col() +
  geom_col(data = prepost_neg) +
  coord_flip() +
  theme_bw() +
  theme(
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank(),
    text = element_text(size = 15),
    strip.text.y = element_text(angle = 0)
  ) +
  xlab("") +
  ylab("Percentage") +
  scale_y_continuous(limits = c(-1, 1), breaks = seq(from = -1, to = 1, by = 0.5), labels = c(100, 50, 0, 50, 100)) +
  scale_fill_manual(values = likert_palette) +
  facet_grid(skill ~ Section, labeller = label_wrap_gen(width=10))
```

It is difficult to gather much of a conclusion from this figure. Many categories appear to have made an improvement, while others seem to show a decrease in agreement from the pre- to the post-survey. Additionally, the figure shows overall trends in the sections, and does not utilize the potential for matching pre- and post-responses from the same student to measure change at the individual level.

>>>>

---
title: "Teaching modeling in introductory statistics: A comparison of formula and tidyverse syntaxes"
blinded: 0
authors:
- name: Amelia McNamara ^[amelia.mcnamara@stthomas.edu]
  affiliation: |
    | Department of Computer & Information Sciences, University of St Thomas
  email: amelia.mcnamara@stthomas.edu
keywords: |
  R language, instruction, data science, statistical computing

abstract: >
  This paper reports on a head-to-head comparison run in a pair of introductory statistics labs, one conducted fully in the formula syntax, the other in tidyverse. Analysis of incidental data from YouTube and RStudio Cloud show interesting distinctions. The formula section appeared to watch a larger proportion of pre-lab YouTube videos, but spend less time computing on RStudio Cloud. Conversely, the tidyverse section watched a smaller proportion of the videos and spent more time computing. Analysis of lab materials showed that tidyverse labs tended to be slightly longer in terms of lines in the provided RMarkdown materials, but not in minutes of the associated YouTube videos. The tidyverse labs exposed students to slightly more distinct R functions, but both labs relied on a quite small vocabulary of consistent functions, which can provide a starting point for instructors interested in teaching introductory statistics in R. Analysis of pre- and post-survey data show no differences between the two labs, so students appeared to have a positive experience regardless of section. This work provides additional evidence for instructors looking to choose between syntaxes for introductory statistics teaching. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    keep_tex: yes
    includes:
      in_header: header.tex
    citation_package: natbib
header-includes:
-  \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
---

@@

lablines <- read_csv(here("data", "processed", "lablines.csv"),
  col_types = cols(
    file = col_character(),
    lines = col_double(),
    section = col_character()
  )
) %>%
  mutate(section = if_else(section == "tidy", "tidyverse", section))

@@

In R, the development of less-complex code for students has been under consideration for even longer [@pruimetal2011]. R offers non-standard evaluation, which allows package authors to create new 'syntax' for their packages [@morandatetal2012]. In human language, syntax is the set of rules for how words and sentences should be structured. If you use the wrong syntax in human language, people will hear there is something wrong with how you structured your speech or writing. However, because human understanding is flexible, the listener will probably still understand the general idea you were trying to convey. Syntax in programming languages is more formal-- it governs what code will execute, run, or compile correctly. Using the wrong syntax usually means failing to get a result from the program.

@@

There is some agreement about pedagogical decisions for teaching R. In particular, most educators agree that in order to reduce cognitive load, instructors should only teach one syntax, and to be as consistent as possible about that syntax [@mcnamaraetal2021a]. There is also some agreement base R syntax is not the appropriate choice for introductory statistics, but there is widespread disagreement on whether the formula syntax or `tidyverse` syntax is better for novices.

While there are strongly-held opinions on which syntax should be taught [@pruimetal2017; @cetinkaya-rundeletal2022], there is relatively little empirical evidence to support these opinions. In the realm of computer science, empirical studies by Andreas Stefik, et al have shown significant differences in the intuitiveness of languages, as well as error rates, based on language design choices [@stefiketal2011; @stefiksiebert2013]. Thus, it seems likely there are language choices that could make data science programming easier (or harder) for users, particularly novices.

@@

The remainder of this paper is organized into three sections. Section \ref{sec:methods} describes the setup of the classes, the participants (\ref{sec:participants}) and their experience (\ref{sec:priorexp}), and the content of the course under investigation (\ref{sec:materials}). Section \ref{sec:results} contains results of the analysis, including a comparison of material lengths between the sections (\ref{sec:ll}), the number of unique functions shown in each section (\ref{sec:numfunc}), results from the pre- and post-survey (\ref{sec:prepost}), and analysis of YouTube (\ref{sec:yt}) and RStudio Cloud (\ref{sec:rstudio}) data. Finally, Section \ref{sec:discussion} discusses the results and opportunities for future study.

All pedagogical materials used for the course under discussion are available on GitHub and are Creative Commons licensed, so they can be used or remixed by anyone who wants to use them (\url{https://github.com/AmeliaMN/STAT220-labs}). All code and anonymized data from this paper is also available on GitHub, for reproducibility (\url{https://github.com/AmeliaMN/ComparingSyntaxForModeling}). Data analysis was performed in R, and the paper is written in RMarkdown. The categorical color palette was chosen using Colorgorical [@gramazioetal2017], and colors for the Likert scale plot are from ColorBrewer [@harrowerbrewer2003]. Example data used throughout the paper is from `palmerpenguins` [@horstetal2020]. Code from the formula section uses the R packages loaded in that course, `mosaic` and `ggformula` (`ggformula` is now loaded automatically with `mosaic`) [@pruimetal2017; @kaplanpruim2020]. Code from the tidyverse section uses the functions from that course, the `tidyverse` and `infer` packages  [@wickhametal2019; @brayetal2021].

@@

The author did this head-to-head comparison in her introductory statistics labs. The comparison was run twice, once in the Spring 2020 semester and once in the Fall 2020 semester. The disruption of COVID-19 to the Spring 2020 semester made the resulting data unusable, so this paper focuses on just Fall 2020 data.

Data was collected from YouTube analytics for watch times, from RStudio Cloud for aggregated compute time, and from pre- and post-surveys of students. Participants for the pre- and post-survey were recruited from this pool after Institutional Research Board ethics review (University of St Thomas IRB 1605810-2).

@@

Participants were students enrolled in an introductory statistics course at a mid-sized private university in the upper Midwest. At this university, statistics students enroll in a lecture (approximately 60-90 students per section), which is broken into several smaller lab sections for hands-on work in statistical software. Lecture and lab sections are taught by different instructors, and the lab sections associated with a particular lecture often use different software. For example, one lab may use Minitab while the other two use Excel. However, every lab section (no matter what lecture it is associated with, or what software is used) does the same set of standardized assignments. This structure provides a consistent basis for comparison.

@@

Each week, the lab instructor prepared a "pre-lab" document in RMarkdown. The pre-lab covered the topics necessary to complete the standardized lab assignment done by all students across lab sections. Pre-lab documents included text explanations of statistical and R programming concepts, sample code, and blanks (both in the code and the text) for students to fill in as they worked.

The instructor recorded YouTube videos of herself working through the pre-lab documents for each lab, and posted them in advance. Literature about flipped classrooms suggests shorter videos are better for student engagement, although there is no consensus about the ideal length for videos, with suggestions ranging from 5 to 20 minutes as a maximum length for a video [@zuber2016; @beattyetal2019; @guoetal2014]. The instructor attempted to keep the total number of minutes of video content below 20 each week. If video content became too long, the instructor split the content into multiple shorter videos. Students were told to watch the pre-lab video(s) and work through the RMarkdown document on their own time, then come to synchronous class to ask questions and get help starting on the real lab assignment.

@@

One obvious question arising when considering the comparison of the two syntaxes is whether students performed better in one section or another. The IRB did not cover examining student work (an obvious place for improved further research), so we cannot look at student outcomes on a per-assignment basis. However, running a randomization test for a difference in overall mean lab grades showed no significant difference between the two sections. While they may have been interesting differences in grades depending on the topic of the lab, we at least know these differences averaged out in the end.

@@

The length of the pre-lab RMarkdown documents can be measured using lines. Figure \ref{fig:prelablength} shows the number of lines of code for each section's pre-lab document, per week. Lines in the RMarkdown document include the YAML header (consistent between documents), the descriptive text about processes (largely similar between documents), and the code in code chunks, which varies based on the syntax of the lab.

@@

```{r}
labsfortesting <- lablines %>%
  drop_na(file) %>%
  pivot_wider(week, names_from = section, values_from = lines) %>%
  mutate(diff = tidyverse - formula)
```

```{r, eval = FALSE}
qqnorm(labsfortesting$diff)
qqline(labsfortesting$diff)
```

```{r}
labttest <- tidy(t.test(labsfortesting$diff))
library(infer)
set.seed(11082021)
labboots <- labsfortesting %>%
  specify(response = diff) %>%
  generate(reps = 5000, type = "bootstrap") %>%
  calculate(stat = "mean")

labboot <- labboots %>%
  get_ci()
```

```{r, eval = FALSE}
qqnorm(labboots$stat)
qqline(labboots$stat)
```

@@

youtube_videos %>%
  mutate(week = case_when(section == "formula" ~  week - 0.15,
                          section == "tidyverse" ~ week + 0.25)) %>%
  ggplot(aes(x = week, y = tot_min, fill = section)) +
  geom_col(position = "stack", color = "white", width = 0.4) +
  scale_x_continuous(breaks = seq(from = , to = 15, by = 2)) +
  # facet_wrap(~section, nrow = 2) +
  ylab("Total length of pre-lab \n videos (minutes)") +
  xlab("Week of semester") +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.title.y = element_text(size = 14)
  )
```

```{r}
youtube_fortesting <- youtube_weeks %>%
  select(week, section, tot_min) %>%
  pivot_wider(week, names_from = section, values_from = tot_min) %>%
  mutate(diff = tidyverse - formula, 
         avglength = (tidyverse + formula)/2, 
         perc_diff = abs(tidyverse - formula)/avglength)
```

```{r, eval = FALSE}
qqnorm(youtube_fortesting$diff)
qqline(youtube_fortesting$diff)
```

```{r}
yttest <- tidy(t.test(youtube_fortesting$diff))
ytboots <- youtube_fortesting %>%
  specify(response = diff) %>%
  generate(reps = 5000, type = "bootstrap") %>%
  calculate(stat = "mean") 
ytboot <- ytboots %>%
  get_ci(point_estimate = mean(youtube_fortesting$diff), type = "bias-corrected")
detach("package:infer")
```

```{r, eval = FALSE}
qqnorm(ytboots$stat)
qqline(ytboots$stat)
```

```{r, eval = FALSE}
linesvminutes <- lablines %>%
  left_join(youtube_videos, by = c("week", "section"))
m1 <- lm(tot_min~lines, data = linesvminutes)
m2 <- lm(tot_min~lines+section, data = linesvminutes)
m3 <- lm(tot_min~lines*section, data = linesvminutes)
anova(m1, m2, m3)
summary(m1)

ggplot(linesvminutes, aes(x=lines, y=tot_min, color = section)) + geom_point() 
```

Every attempt was made to align these RMarkdown documents, so the descriptive text was only changed when necessary to describe specific elements of the code. Similarly, if blank code chunks appeared in one lab, that was mirrored by a blank chunk in the other lab. Both labs' documents were styled using the styler package [@mullerwalthert2022] to remove inconsistencies with spacing, assignment operators, and the like. Code was styled using the default style from the package, which is based on the tidyverse style guide [@wickham2022]. The tidyverse style guide has become the de facto style guide for R, as previously-existing style guides like the Google R Style Guide have largely decided to follow @wickham2022.

Figure \ref{fig:prelablength} indicates RMarkdown documents for the `tidyverse` section tended to be longer. We can compute a difference in lab lengths for each week, and compute the mean difference, which is `r round(mean(labsfortesting$diff))` lines. Although we only have 12 labs worth of data, it appeared relatively normally distributed. A 95\% confidence interval for the difference in line length (computed using the Student's t-distribution) is (`r round(labttest$conf.low)`, `r round(labttest$conf.high)`), and a bootstrap confidence interval computed with 5,000 bootstrap samples and using the percentile method is (`r round(labboot$lower_ci)`, `r round(labboot$upper_ci)`). Both intervals indicate labs for the `tidyverse` section were longer, but only by a few lines.

A slightly longer length for these labs makes sense, because `tidyverse` code is characterized by multiple short lines strung together into a pipeline with `%>%`, while the formula syntax typically has single function calls, sometimes with more arguments. The code shown in this paper is styled the same way the labs were, so any code comparison you see (for example, in \ref{tally-ex1} and \ref{tidy-tally1}) will show the difference in lines of code between the two syntaxes.

@@

Again, we can compute a pairwise difference in total video length (adding together multiple videos in weeks that had them), and compute the mean of that difference. That difference is `r round(mean(youtube_fortesting$diff))` minutes (`tidyverse` videos being longer). The distribution of differences appeared to be right-skewed, which may impact results because the sample size is so low. Using a t-distribution, we can compute a 95\% confidence interval, which is (`r round(yttest$conf.low, 1)`, `r round(yttest$conf.high, 1)`), and thus contains 0.

Alternately, we could compute a bias-corrected 95\% bootstrap confidence interval using 5,000 bootstrap samples. Because of the right-skew to the data, the boostrap distribution also appeared skewed. Bias-correction should help adjust for this. The boostrap interval is (`r round(ytboot$lower_ci, digits=2)`, `r round(ytboot$upper_ci)`). This interval does not contain 0, but `r round(ytboot$lower_ci, digits=2)` minutes is equivalent to `r round(ytboot$lower_ci*60, digits=0)` seconds, which is not practically significant. It appears that while `tidyverse` labs are longer in terms of lines of code, the corresponding videos are not meaningfully different in length.

@@

A similar discrepancy can be seen in week 10, where the formula section's RMarkdown document was `r pull(filter(lablines, week == 10, section == "formula"), lines)` lines long, and the videos totaled `r round(filter(youtube_weeks, week == 10, section == "formula") %>% pull(tot_min))` minutes. That same week, the `tidyverse` RMarkdown document was `r pull(filter(lablines, week == 10, section == "tidyverse"), lines)` lines long, and the videos totaled `r round(filter(youtube_weeks, week == 10, section == "tidyverse") %>% pull(tot_min))` minutes.

@@

These discrepancies made it take longer to explain code in the `tidyverse` section for these topics. However, as we saw in \ref{sec:ll}, overall the length of videos did not appear meaningfully different between sections. Comparisons of RMarkdown document length and YouTube video length, as well as the corresponding reasons for those discrepancies are the first hint of the computing time results to come in Section \ref{sec:rstudio}.

@@

Of course, while teaching a real class, an instructor often has to improvise at least a little. So, it is also interesting to consider the number of functions actually shown throughout the course of the semester. To do this, we can consider the functions shown in the filled-in version of pre-lab documents the instructor ended up with after recording the associated instructional video.

@@

The functions both sections of students saw included helper functions like `library()`, `set.seed()`, and `set()` (a function in the knitr options included in the top of each RMarkdown document), statistics like `mean()`, `sd()`, and `cor()`, and modeling-related functions like `aov()`, `lm()`, `summary()` and `predict()`. (The instructor chose not to introduce the `tidymodels` package to the `tidyverse` section.)

@@

Students in the formula section saw `r nrow(formulafunctions)` functions unique from the set both sections saw, while the `tidyverse` section saw `r nrow(tidyfunctions)` unique functions. It makes sense the number of unique functions in the `tidyverse` section would be slightly larger, as there are several elements of the `tidyverse` that require multiple functions to combine to accomplish a single task.

For example, students in both sections saw how to make a barchart, boxplot, histogram, and scatterplot, but in the formula section they used standalone functions like `gf_boxplot()` whereas in the `tidyverse` section they needed to start with `ggplot` and add on a `geom_` function like `geom_boxplot()`, while specifying the `aes`thetic values somewhere.

Similarly, both sections saw several common summary statistics, but in the formula section they used the function (e.g. `mean()`) on its own, whereas in the `tidyverse` section base summary functions needed to be wrapped within `summarize()`. Students in the `tidyverse` section also saw slightly more summary statistic functions, because one lab called for the five number summary.

@@

In the `tidyverse` section, the instructor chose to show two approaches. Both approaches are in \ref{tidy-summary1}, and both needed to include `drop_na()` to deal with missing values. Past those similarities, the approaches are divergent.

@@

The instructor should have chosen a single solution to present to students, but was faced with a dilemma. The first `tidyverse` approach is very verbose, but it follows nicely from other summary statistics students had already seen, just adding a few more functions like `min`, `max`, and `quantile`. The second solution is more concise, but it introduces the `pull` function, which was never used again in the course. 
This brings up an important consideration when teaching coding-- how many times students will see the same function. Because there is some cognitive load associated with learning a new function, and repetition helps move information from working memory to long term memory, it is ideal for students to see each function at least twice [@lovettgreenhouse2000; @mcnamaraetal2021a]. When analyzing the number of functions shown in each section, we found there were `r dim(filter(allfunctions, n == 1, section == "formula"))[1]` functions shown only one time in the formula section, and `r dim(filter(allfunctions, n == 1, section == "tidyverse"))[1]` functions only shown once in the `tidyverse` section.

Overall, neither section appeared to expose students to an overwhelming number of functions. One argument against the use of the  `tidyverse` in teaching is that it contains too many functions. However, when teaching a course (particularly an introductory one) an instructor never shows every function in a package. The `tidyverse` section saw `r nrow(tidyfunctions)` as compared to the formula section's `r nrow(formulafunctions)` functions, but that difference does not feel practically significant, particularly considering the way in which `tidyverse` operations often use helper functions.

The comparison also underscores the fact that while instructors may say they are teaching `tidyverse` or formula syntax, they are ultimately teaching R. Both sections saw `r nrow(bothfunctions)` common functions, many of them from base R.

The practice of analyzing the number of functions shown over the course of the semester was eye-opening. It will provide valuable information for the instructor the next time she teaches the course, as she can attempt to remove functions only shown once, and ensure the cheatsheets better match what is actually shown throughout the semester.  The list of functions provided in Appendix \ref{sec:functions} can also serve as a starting point for other instructors as they work to produce curricular materials for introductory statistics classes in R.

@@

As discussed in \ref{sec:participants}, the number of students who completed both the pre- and post-surveys were low ($n=8$ for the formula section, and $n=13$ for the `tidyverse` section), so there are major limitations to paired analysis.

@@

\begin{itemize}
\itemsep-3mm
```{r, results = 'asis'}
qs %>%
  select(Q11.x:Q16.x) %>%
  pivot_longer(cols = everything(), names_to = "old", values_to = "q") %>%
  select(-old) %>%
  mutate(q = paste("\\item", q, "\n")) %>%
  pull(q) %>%
  walk(cat)
```
\end{itemize}

In Figure \ref{fig:pre-post}, you can see a visualization of these Likert-scale questions. Based on results below, it became clear that there was no major difference between the two sections, so this plot shows the overall trends without breaking into the groups.

```{r}
pre_sum <- prepostsurvey %>%
  select(Q11.x:Q16.x) %>%
  pivot_longer(everything(),
    names_to = "skill",
    values_to = "feeling"
  )
pre_sum <- pre_sum %>%
  drop_na(feeling) %>%
  group_by(skill, feeling) %>%
  summarize(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  select(-n) %>%
  mutate(type = "pre")

post_sum <- prepostsurvey %>%
  select(Q11.y:Q16.y) %>%
  pivot_longer(everything(),
    names_to = "skill",
    values_to = "feeling"
  )
post_sum <- post_sum %>%
  drop_na(feeling) %>%
  group_by(skill, feeling) %>%
  summarize(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  select(-n) %>%
  mutate(type = "post")

@@

```{r pre-post, fig.cap = "Pre and post responses to Likert-scale questions. Most questions show some level of improvement, such as the first question, `I am confident in my ability to make use of programming software to work with data.' but others show no change or even a decline in agreement.", out.width = "95%", fig.asp = 0.618}
likert_palette <- c("1 - strongly disagree" = "#a6611a", "2" = "#dfc27d", "3 - neutral" = "grey90", "4" = "#80cdc1", "5 - strongly agree" = "#018571")
ggplot(prepost_pos, aes(x = type, y = prop, fill = feeling)) +
  geom_col() +
  geom_col(data = prepost_neg) +
  coord_flip() +
  theme_bw() +
  theme(
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank(),
    text = element_text(size = 15),
    strip.text.y = element_text(angle = 0)
  ) +
  xlab("") +
  ylab("Percentage") +
  scale_y_continuous(limits = c(-1, 1), breaks = seq(from = -1, to = 1, by = 0.5), labels = c(100, 50, 0, 50, 100)) +
  scale_fill_manual(values = likert_palette) +
  facet_grid(rows = vars(skill), labeller = label_wrap_gen(width=10))
```

Figure \ref{fig:pre-post} does not suggest that there was any overall trend from pre- to post-survey. Many categories appear to have made an improvement (like Programming Confident), while others seem to show a decrease in agreement from the pre- to the post-survey (like Overcome Problem).
